---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# generations

Convert birth years to generation names.

This package contains functions that will convert years of birth into names of generational cohorts. For example, `1989` becomes `"Millennial"`. The package contains several helper functions to modify the underlying lookup table as well as way to query it.

## Installation

The package currently lives on GitHub, so you can install it like you would with any other package on GitHub:

```{r, eval = FALSE}
remotes::install_github("joeystanley/generations")
```

You can then load it like you can with any library.

```{r}
library(generations)
```

For the purposes of this tutorial, I'll load `ggplot` as well.

```{r, message = FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)
```

## Converting years to generations

The main function in this package is `generations`. Given a vector of integers, it'll return a factor of generation names.  First, I'll generate some random years of birth.

```{r, echo = FALSE}
set.seed(12)
```


```{r}
yobs <- floor(runif(10, 1900, 2020))
yobs
```

I can now easy convert that into generations.

```{r}
generations(yobs)
```

This function works on any year betwen 1435 and 2030. Numbers outside that range return `NA`. 

Note that by default, the function will return the vector as *factor*, with the levels ordered so that the oldest generation in the vector is first. To get a character vector instead, add the argument `as_factor = FALSE`.

### Customizing output

There are some tweaks you can do to adjust the output of `generations`. First, you can return longer forms of the generational names by specifying `full_names = TRUE`. 

```{r}
generations(yobs, full_names = TRUE)
```

What this does is simply add `"Generation"` to the end of each one, unless it's `"Gen X"` (or Y, or Z), in which case it'll expand it out to simply `"Generation X"`.

You can also show the years included in each generation by adding the `years = TRUE` argument. This will add a space and, inside a pair of parentheses, the start and end years of that generation, separated by an en dash. 

```{r}
generations(yobs, years = TRUE)
```

The primary purpose of this is for visualizations, since not everyone is familiar with (or agrees with) the year ranges.

```{r, echo = FALSE}
set.seed(10)
```

```{r}
many_yobs <- tibble(yob = floor(rnorm(1000, 1975, 15))) %>%
  mutate(gen = generations(yob, full_names = TRUE, years = TRUE))
ggplot(many_yobs, aes(yob, fill = gen)) + 
  geom_histogram(binwidth = 1) + 
  scale_fill_brewer(name = NULL, palette = "Set1")
```

How this additional portion is formatted can be adjusted. If rendering an en dash is troublesome for you, you can change it to something else with `years_range_sep`. You may also want to change the space between the generation name and the opening parenthesis into a newline character with `years_sep`, again for visualization purposes.

```{r}
many_yobs <- many_yobs %>%
  mutate(gen = generations(yob, full_names = TRUE, years = TRUE,
                           years_sep = "\n", years_range_sep = " to "))

ggplot(many_yobs, aes(yob, fill = gen)) + 
  geom_histogram(binwidth = 1) + 
  scale_fill_brewer(name = NULL, palette = "Set1") + 
  labs(x = NULL) + 
  theme(legend.key.height = unit(1, "cm"))
```

If you want to get really fancy, you can make the legend keys approximate the width they take up on the *x*-axis and put better tics marks.

```{r, message = FALSE, warning = FALSE}
widths <- many_yobs %>%
  group_by(gen) %>%
  summarize(width = max(yob) - min(yob)) %>%
  ungroup() %>%
  mutate(width = width / max(width) * 1.4) # you may have to fudge this a little more

ggplot(many_yobs, aes(yob, fill = gen)) + 
  geom_histogram(binwidth = 1) + 
  scale_fill_brewer(name = NULL, palette = "Set1") + 
  scale_x_continuous(breaks = c(1929, 1946, 1964, 1984, 2008, 2030)) + 
  labs(x = NULL) + 
  theme(legend.position = "bottom") + 
  guides(fill = guide_legend(nrow = 1, label.position = "bottom", 
                             keywidth = widths$width, default.unit = "inches"))
```

## Querying generation data

To see a list of the generational data, you can use `show_generations`, which will return a data frame containing the names, start years, and end years.

```{r}
show_generations()
```

You can also get simple information. For example, if you want to know when the start or end year of a particular generation is, you can use `get_start()` or `get_end()`:

```{r}
get_start("Silent")
get_end("Millennial")
```

You can also find the names of neighboring generations with `get_prev_gen()` and `get_next_gen()`. 

```{r}
get_next_gen("Millennial")
get_prev_gen("Missionary")
```

Note that if ask for something newer than Gen Z or older than Aurthurian it will return `NA`.

## Customizing generation data

The data that this package uses is loaded as a hidden object when you load the package. You may modify it with the functions described in this section. These changes will affect the dataset so long as the `generations` package is loaded. You'll have to reset the data each time to reload it. 

The labels and years for each generation are mostly borrowed from Howe & Strauss' Generational Theory books. However, not everyone agrees on the names and year ranges for the various generations. For this reason, the generations package makes it easy to modify the generations data to your liking. 

To rename a generation, use `rename_generation()`, with the old name first and the new name second. For example, if you want to use *Zoomer* instead of *Gen Z*, you can do so.

```{r}
rename_generation("Gen Z", "Zoomer")
```

You'll get a message informing you that the change has been made. If you now run `show_generations()` you'll see that the change has been made and if you rerun `generations()`, you'll get updated results.

```{r}
show_generations()
generations(yobs)
```

Because many people may want to use the term *Zoomer* instead of *Gen Z*, a shortcut function, `use_zoomer`, which is just a wrapper around `rename_generation("Gen Z", "Zoomer")`, is included in the package. The other shortcut functions are `use_gen_y()`, `use_13th()`, `use_baby_boom()` as well as their reciprocals `use_gen_z()`, `use_millennial()`, `use_gen_x()` and `use_boomer()`. 

You may also want to change the years. For example, many people consider 1997 as the end of the Millennial Generation. You can make this change with `redefine_generation`. With this function, you must specify the new start and the new end year.

```{r}
redefine_generation("Millennial", 1983, 1997)
```

Since changing one generation impacts adjacent generations, you'll get a message showing you what the new ranges are for this, the previous, and the next generations.

You can reset the data back to its original form with `reset_generations()`.



